Thesis: draft

### 2021-05-11 draft.tex
made small changes

\epigraph{\itshape At times, our own light goes out and is rekindled by a spark from another person. Each of us has cause to think with deep gratitude of those who have lighted the flame within us.}{---Albert Schweitzer}
\vspace*{\fill}
\noindent
First and foremost, I would like to thank my supervisor Prof. Luis E.C. Rocha for his support, feedback and useful insights during the course of this year. This thesis would not be of the same quality if it wasn't for his insightful remarks. I learned a tremendous deal, not only about social physics and network science, but also about organizing, structuring and reporting relevant results. I would not have been able to learn these things without his guidance.\\
\newline
I also want to give special thanks to my co-supervisor Prof. Jan Ryckebusch. Not only for his support during my thesis, but mostly for the enthusiastic and passionate way in which he teaches all his classes. His dedication to his students and his willing to learn them valuable insights in the wonderful world of physics are truly admirable.\\
\newline
I am especially thankful to Yolan. He helped me with every possible problem I encountered during my physics education. If it was not for him, I would still be a (very poor) programming beginner. His patience and talent for explaining things in an understandable way are incredibly valuable. I do not only want to thank him for all his help during my study and my thesis, but also for being the great friend he is. This thesis, but also myself, would not have been the same without him.\\
\newline
Beside Yolan, I want to give credit to all of my other friends. I would not be the same person without any of them.\\
\newline
Finally, I would like to thank my family, and especially my parents. They have unbelievable faith in me and support me in every single way. I feel great gratitude for how they raised me and for all the opportunities that they have given me, and are still giving me every single day.\\
\vfill
\noindent
Nina Botte.

@@

Door het toenemende gebruik van sociale media worden deze echter overspoeld door informatie. Om hun gebruikers niet het gevoel te geven dat ze verdrinken in wat ze te zien krijgen op hun sociale media platform, gebruiken deze sociale media platformen  filtering algoritmen. Deze filtering algoritmen filteren, zoals de naam al impliceert, en individualiseren de informatie die iemand te zien krijgt op zijn/haar tijdslijn. Een gebruiker op sociale media krijgt dus niet alle berichten die de vrienden posten te zien, maar enkel een selectie die onder andere gemaakt wordt op basis van de gebruiker zijn/haar sociale media geschiedenis. Dit filteren/individualiseren kan, mogelijks, grote gevolgen hebben. Zo zou het kunnen leiden tot de vorming van  ``echo kamers'' waar gebruikers gevangen raken tussen ideeën die hun eigen visie versterken. Dit kan op zijn beurt polarisatie in de maatschappij, opkomst van extremisme, enzovoort versterken. Het is daarom uiterst belangrijk om de impact van deze filtering algoritmen grondig in kaart te brengen.\\

@@

One of the goals of this thesis is to investigate/determine the role of the underlying network structure on the formation and dynamics of opinions in the system. Some examples of theoretical network models used in this thesis are the Erd\H{o}s-R\'{e}nyi network, the stochastic block model, the Watts-Strogatz model, etc. These will be explained thoroughly in Chapter \ref{chap2}, Section \ref{netModel}.

@@

As stated before, the formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc.  Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
\newline
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
\newline
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. A small overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
\newline
One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of nodes is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in  this thesis will be given.

@@

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals would be well know, then the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. The use of statistical physics may, however, also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in regular physical research and science in general.

@@

The adjacency matrix contains a lot of information such as the degree of a node. The degree $k_i$ of a node $i$ is the number of edges attached to node $i$ or, in others words, $k_i$ represents the number of nearest neighbors of node $i$. The degree can be easily obtained from the adjacency matrix \cite{Mata2020}
\begin{equation}
	k_i = \sum_{j=1}^N A_{ij}\ .
\end{equation}
In the case of a directed graph one can differentiate between the number of incoming edges $k_i^{\textrm{in}}$ and the number of outgoing edges $k_i^{\textrm{out}}$ of node $i$. These are called the in-degree and out-degree respectively and are defined as \cite{Mata2020}
\begin{align}
	k_i^{\text{in}} = \sum_{j=1}^N A_{ji} \ , && k_i^{\text{out}} = \sum_{j=1}^N A_{ij} \ .
\end{align}
The total degree of node $i$ is then $k_i = k_i^{\text{in}} + k_i^{\text{out}}$. For weighted graphs the degree is easily generalized to the weighted degree $s_i$, often called strength \cite{Ioannis2007}
\begin{equation}
	s_i = \sum_{j=1}^N w_{ij} \ .
\end{equation}
From the degree of the nodes in the network one can construct the degree distribution. The degree distribution $P(k)$ represents the fraction of nodes with a degree $k$ or, in other words, the degree distribution gives the probability that a randomly selected node has a degree $k$ \cite{Newman2003}. The average degree can be obtained from the degree distribution \cite{Mata2020}
\begin{equation}
	\left<k\right> = \frac{1}{N} \sum_{i=1}^N k_i = \sum_k k P(k) \ ,
\end{equation}
where $N$ is the total number of nodes in the network. The degree distribution also allows us to classify networks. The two most important classes are homogeneous and heterogeneous networks. Homogeneous networks have a bell curved degree distribution, eg. a Poisson distribution. In this case, most of the nodes have a degree close to the average degree $\left<k\right>$ \cite{Barabasi2016}. Heterogeneous networks, on the other hand, have a power-law degree distribution, $P(k) \sim k^{-\gamma}$. They are also referred to as scale-free networks, since they do not posses a characteristic length scale or, in this case, the average degree is not a characteristic scale for the network \cite{Barabasi2016}\cite{Mata2020}.

@@

Another concept that is closely related to the degree of a node is the notion of assortative and disassortative networks. More precisely, the notion of assortativity and disassortativity has to do with degree correlations. \\
\newline
The joint degree distribution $P(k, k')$ is the probability that an arbitrary edge connects a node with degree $k$ to a node with degree $k'$ \cite{F.Costa2007}. The conditional probability $P(k|k')$, which is another way to express dependencies between node degrees, is the probability that an arbitrary neighbor of a node with degree $k$ has a degree $k'$ \cite{F.Costa2007}
\begin{equation}
	P(k|k') = \frac{\left<k\right>P(k, k')}{kP(k)} \ .
\end{equation}
The average degree of the nearest neighbors of nodes with degree $k$ can be computed as \cite{F.Costa2007}
\begin{equation}
	k_{nn}(k) = \sum_{k'} k'P(k'|k) \ .
\end{equation}
If there are no degree correlations, $k_{nn}$ is independent of $k$. Such a network is called a neutral network \cite{F.Costa2007}. If $k_{nn}(k)$ is an increasing function of $k$, then nodes with a high degree tend to connect to nodes with a high degree and the network is classified as assortative. If instead $k_{nn}(k)$ is a decreasing function of $k$, nodes with a high degree tend to connect to nodes with a low degree and the network is called disassortative \cite{F.Costa2007}.\\
\newline
The terms assortativity and disassortativity are also often used without taking degree correlations into account. Instead, one may look at a particular node property to classify networks as either assortative or disassortative \cite{Thed2014}. In this way, an assortative network is a network where nodes tend to connect to other nodes with similar properties as themselves, whereas in a disassortative network nodes tend to connect to nodes with different properties \cite{Thed2014}.

@@

Several useful measurements quantify the notion of length and distance in a network/graph. We start with the definition of a path: a path is a sequence of distinct nodes, such that adjacent nodes in the sequence are adjacent nodes in the network \cite{Goddard2010}. The length of a path is then defined as the number of edges in the path \cite{Goddard2010}. A geodesic path, often called the shortest path, between two nodes is the path between two nodes $i$ and $j$ with minimum length (the geodesic path is not necessarily unique) \cite{F.Costa2007}. The length of the geodesic path between the two nodes $i$ and $j$ is called the geodesic distance or shortest distance $d_{ij}$ \cite{F.Costa2007}. Geodesic distance is often just called distance \cite{Goddard2010}. The notion of distance enables to define several global network measurements such as average distance, diameter and radius.\\  
\newline
\textbf{Average distance}\\
\newline
The average distance is the mean value of the geodesic distance $d_{ij}$ \cite{F.Costa2007}
\begin{equation}\label{avdist}
	l = \frac{1}{N(N-1)} \sum_{i \neq j} d_{ij} \ .
\end{equation}
This definition, however, diverges if there are unconnected nodes in the network (unconnected nodes have, by definition, a distance equal to infinity) \cite{F.Costa2007}. This problem could be avoided by only including connected pairs in the sum, but this introduces a distortion for networks with a lot of unconnected pairs of nodes. These networks will then have a small average distance, which is only expected for networks with lots of connections \cite{F.Costa2007}. Instead, another definition of average distance is used: the global efficiency $E$ \cite{F.Costa2007}
\begin{equation}
	E = \frac{1}{N(N-1)} \sum_{i \neq j} \frac{1}{d_{ij}} \ .
\end{equation}
This measurement quantifies the efficiency of the network in sending information between nodes \cite{F.Costa2007}. This assumes that the efficiency of sending information between two nodes is reciprocal to their distance. The harmonic mean of the geodesic distances is now defined as the reciprocal of the global efficiency \cite{F.Costa2007}
\begin{equation}
 	h = \frac{1}{E} \ .
\end{equation}
\newline
\textbf{Diameter and radius}\\
\newline
The diameter $diam(G)$ of a network $G$ is the maximum distance between any pair of nodes \cite{Goddard2010}. The eccentricity of a node is the maximum distance from that node to any other node. The radius $rad(G)$ is the minimum eccentricity among all nodes of $G$. Note that the diameter is the maximum eccentricity among all nodes in $G$ \cite{Goddard2010}.
\newpage
\noindent
\subsection{Clustering coefficient and transitivity}

@@

Let us denote the number of triangles as $N_{\Delta}$ and the number of connected triples as $N_3$. These two numbers can be obtained from the adjacency matrix in the following way \cite{F.Costa2007}

@@

It is not too hard to see that $N_{\Delta}(i)$ counts the number of edges between the neighbors of $i$ and that $N_3(i)$ is equal to $k_i(k_i-1)/2$, where $k_i$ is the degree of node $i$. It is then obvious that the local clustering coefficient $cc_i$ (Eq. (\ref{clus})) of a node and the transitivity $T_i$ (Eq. (\ref{trans})) of a node define the same quantity \cite{F.Costa2007}. In the case of the average clustering coefficient $\left<cc\right>$ and the global transitivity $T$ this is, however, not the case. The difference between the two definitions is that the average of Equation (\ref{globalTrans}) gives the same weight to each triangle in the network, whereas Equation (\ref{avClus}) gives the same weight to each node \cite{Newman2003}. This may lead to slightly different values since nodes with a higher degree may possibly be involved in a higher number of triangles than nodes with a lower degree \cite{F.Costa2007}.

@@

The modularity for a network with $N$ nodes and $L$ edges and partitioned into $n_c$ communities is calculated as follows. Each community has $N_c$ nodes and $L_c$ edges ($c = 1,..., n_c$) \cite{Albert2016}. The modularity of a community $Q_c$ is then calculated by measuring the difference between the network's real wiring diagram (which is given by the adjacency matrix $A_{ij}$) and the expected number of edges between $i$ and $j$ if the network is randomly wired (where $i$ and $j$ are two nodes that lie in the same community $C_c$) \cite{Albert2016}. This last quantity is given by $p_{ij}$, which is obtained by randomizing the original network while keeping the expected degree of each node unchanged \cite{Albert2016}. Hence, the modularity of a community $Q_c$ becomes \cite{Albert2016}

@@

where the factor $\frac{1}{2L}$ is just a normalization factor. Using \cite{Albert2016}

@@

As said before, modularity is often used in community detection algorithms. It is thus important to be aware of the limitations of modularity. One of the biggest shortcomings of modularity maximization as community detection is its resolution limit. Modularity maximization forces small communities into larger ones \cite{Albert2016}. If two communities $A$ and $B$ are merged, the modularity changes by \cite{Albert2016}

@@

and if there is at least one edge between the two communities; even if $A$ and $B$ are distinct communities \cite{Albert2016}! Equation (\ref{reslim}) is called the resolution limit: communities that are smaller than this limit will not be detected by modularity maximization \cite{Albert2016}.

@@

	p_k = \binom{N-1}{k}p^k (1-p)^{N-1-k} \ .

@@

	\left<k\right> &= \sum_{k=0}^{N-1} k\  p_k  \\

@@

	p_k = \mathrm{e}^{-\left<k\right>} \frac{\left<k\right>^k}{k!} \ .

@@

This can be easily understood since choosing three nodes out of $N$ is given by $\binom{N}{3}$. The chance that they are connected by three edges is given by $p^3$, which gives the number of triangles; the chance that the are connected by two edges is $p^2$, which determines the number of connected triples (recall that this model creates random networks with no self-loops and no edge duplicates). If the average degree $\left<k\right>$ is constant, the average clustering constant goes as $\left<cc\right> \sim \mathcal{O}(1/N)$, which becomes very small for large networks \cite{Clauset2011}. \\

@@

A more precise calculation can be found in Appendix \ref{ER}, Section \ref{ERdist}. Equation (\ref{diameter}) might, however, be a better approximation to the average distance $\left<d\right>$ between two randomly chosen nodes \cite{Albert2014}. This is because $d_{\textrm{max}}$ is often dominated by a few extreme paths, where $\left<d\right>$ instead is averaged over all node pairs and thus suppresses these fluctuations \cite{Albert2014}. Hence, the average distance of a random network is defined by \cite{Albert2014}

@@

To summarize: the random Erd\H{o}s-R\'{e}nyi network has a bell shaped degree distribution which approximates a Poisson distribution for $N \gg 1$; the average clustering coefficient becomes small for large networks and the average distance depends logarithmic on the system sizes. The Erd\H{o}s-R\'{e}nyi network thus possesses one of the two properties of the ``small world phenomenon'' (namely short average distance).\\

>>>>

\epigraph{\itshape At times, our own light goes out and is rekindled by a spark from another person. Each of us has cause to think with deep gratitude of those who have lighted the flame within us.}{---Albert Schweitzer}
\vspace*{\fill}
\noindent
First and foremost, I would like to thank my supervisor Prof. Luis E.C. Rocha for his support, feedback and useful insights during the course of this year. This thesis would not be of the same quality if it wasn't for his insightful remarks. I learned a tremendous deal, not only about social physics and network science, but also about organizing, structuring and reporting relevant results. I would not have been able to learn these things without his guidance.\\
\newline
I also want to give special thanks to my co-supervisor Prof. Jan Ryckebusch. Not only for his support during my thesis, but mostly for the enthusiastic and passionate way in which he teaches all his classes. His dedication to his students and his willing to learn them valuable insights in the wonderful world of physics are truly admirable.\\
\newline
I am especially thankful to Yolan. He helped me with every possible problem I encountered during my physics education. If it was not for him, I would still be a (very poor) programming beginner. His patience and talent for explaining things in an understandable way are incredibly valuable. I do not only want to thank him for all his help during my study and my thesis, but also for being the great friend he is. This thesis, but also myself, would not have been the same without him.\\
\newline
Beside Yolan, I want to give credit to all of my other friends. They make my life valuable, joyful and are a true enrichment.\\
\newline
Finally, I would like to thank my family, and especially my parents. They have unbelievable faith in me and support me in every single way. I feel great gratitude for how they raised me and for all the opportunities that they have given me, and are still giving me every single day.\\
\vfill
\noindent
Nina Botte.

@@

Door het toenemende gebruik van sociale media worden deze echter overspoeld door informatie. Om hun gebruikers niet het gevoel te geven dat ze verdrinken in wat ze te zien krijgen op hun sociale media platform, gebruiken deze sociale media platformen filtering algoritmen. Deze filtering algoritmen filteren, zoals de naam al impliceert, en individualiseren de informatie die iemand te zien krijgt op zijn/haar tijdslijn. Een gebruiker op sociale media krijgt dus niet alle berichten die zijn/haar vrienden posten te zien, maar enkel een selectie die onder andere gemaakt wordt op basis van de gebruiker zijn/haar sociale media geschiedenis. Dit filteren/individualiseren kan, mogelijks, grote gevolgen hebben. Zo zou het kunnen leiden tot de vorming van ``echo kamers'' waar gebruikers gevangen raken tussen ideeën die hun eigen visie versterken. Dit kan op zijn beurt polarisatie in de maatschappij, opkomst van extremisme, enzovoort bevorderen. Het is daarom uiterst belangrijk om de impact van deze filtering algoritmen grondig in kaart te brengen.\\

@@

Some examples of theoretical network models used in this thesis are the Erd\H{o}s-R\'{e}nyi network, the stochastic block model and the Watts-Strogatz model. These will be explained thoroughly in Chapter \ref{chap2}, Section \ref{netModel}.

@@

The formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc. Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
\newline
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
\newline
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. A small overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
\newline
One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of nodes is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in this thesis will be given.

@@

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals would be well know, the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. The use of statistical physics may, however, also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}\cite{Gug2015}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality \cite{Gug2015}. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in regular physical research and science in general.

@@

The adjacency matrix contains a lot of information such as the degree of a node. The degree $k_i$ of a node $i$ is the number of edges attached to node $i$ or, in others words, $k_i$ represents the number of nearest neighbors of node $i$. The degree can be easily obtained from the adjacency matrix \cite{Mata2020}
\begin{equation}
	k_i = \sum_{j=1}^N A_{ij}\ .
\end{equation}
In the case of a directed graph one can differentiate between the number of incoming edges $k_i^{\textrm{in}}$ and the number of outgoing edges $k_i^{\textrm{out}}$ of node $i$. These are called the in-degree and out-degree respectively and are defined as \cite{Mata2020}
\begin{align}
	k_i^{\text{in}} = \sum_{j=1}^N A_{ji} \ , && k_i^{\text{out}} = \sum_{j=1}^N A_{ij} \ .
\end{align}
The total degree of node $i$ is then $k_i = k_i^{\text{in}} + k_i^{\text{out}}$. For weighted graphs the degree is easily generalized to the weighted degree $s_i$, often called strength \cite{Ioannis2007}
\begin{equation}
	s_i = \sum_{j=1}^N w_{ij} \ .
\end{equation}
From the degree of the nodes in the network one can construct the degree distribution. The degree distribution $P(k)$ represents the fraction of nodes with a degree $k$ or, in other words, the degree distribution gives the probability that a randomly selected node has a degree $k$ \cite{Newman2003}. The average degree can be obtained from the degree distribution \cite{Mata2020}
\begin{equation}
	\left<k\right> = \frac{1}{N} \sum_{i=1}^N k_i = \sum_k k P(k) \ ,
\end{equation}
where $N$ is the total number of nodes in the network. The degree distribution allows us to classify networks. The two most important classes are homogeneous and heterogeneous networks. Homogeneous networks have a bell curved degree distribution, eg. a Poisson distribution. In this case, most of the nodes have a degree close to the average degree $\left<k\right>$ \cite{Barabasi2016}. Heterogeneous networks, on the other hand, have a power-law degree distribution, $P(k) \sim k^{-\gamma}$. They are also referred to as scale-free networks, since they do not posses a characteristic length scale or, in this case, the average degree is not a characteristic scale for the network \cite{Barabasi2016}\cite{Mata2020}.

@@

Another concept that is closely related to the degree of a node is the notion of assortative and disassortative networks. More precisely, the notion of assortativity and disassortativity has to do with degree correlations. \\
\newline
The joint degree distribution $P(k, k')$ is the probability that an arbitrary edge connects a node with degree $k$ to a node with degree $k'$ \cite{F.Costa2007}. The conditional probability $P(k|k')$, which is another way to express dependencies between node degrees, is the probability that an arbitrary neighbor of a node with degree $k$ has a degree $k'$ \cite{F.Costa2007}
\begin{equation}
	P(k|k') = \frac{\left<k\right>P(k, k')}{kP(k)} \ .
\end{equation}
The average degree of the nearest neighbors of nodes with degree $k$ can be computed as \cite{F.Costa2007}
\begin{equation}
	k_{nn}(k) = \sum_{k'} k'P(k'|k) \ .
\end{equation}
If there are no degree correlations, $k_{nn}$ is independent of $k$. Such a network is called a neutral network \cite{F.Costa2007}. If $k_{nn}(k)$ is an increasing function of $k$, nodes with a high degree tend to connect to nodes with a high degree and the network is classified as assortative. If instead $k_{nn}(k)$ is a decreasing function of $k$, nodes with a high degree tend to connect to nodes with a low degree and the network is called disassortative \cite{F.Costa2007}.\\
\newline
The terms assortativity and disassortativity are also often used without taking degree correlations into account. Instead, one may look at a particular node property to classify networks as either assortative or disassortative \cite{Thed2014}. In this way, an assortative network is a network where nodes tend to connect to other nodes with similar properties as themselves, whereas in a disassortative network nodes tend to connect to nodes with different properties \cite{Thed2014}.

@@

Several measurements quantify the notion of length and distance in a network/graph. We start with the definition of a path: a path is a sequence of distinct nodes, such that adjacent nodes in the sequence are adjacent nodes in the network \cite{Goddard2010}. The length of a path is defined as the number of edges in the path \cite{Goddard2010}. A geodesic path, often called the shortest path, between two nodes is the path between two nodes $i$ and $j$ with minimum length (the geodesic path is not necessarily unique) \cite{F.Costa2007}. The length of the geodesic path between the two nodes $i$ and $j$ is called the geodesic distance or shortest distance $d_{ij}$ \cite{F.Costa2007}. Geodesic distance is often just called distance \cite{Goddard2010}. The notion of distance enables to define several global network measurements such as average distance, diameter and radius.\\  
\newline
\textbf{Average distance}\\
\newline
The average distance is the mean value of the geodesic distance $d_{ij}$ \cite{F.Costa2007}
\begin{equation}\label{avdist}
	\left<d\right> = \frac{1}{N(N-1)} \sum_{i \neq j} d_{ij} \ .
\end{equation}
This definition, however, diverges if there are unconnected nodes in the network (unconnected nodes have, by definition, a distance equal to infinity) \cite{F.Costa2007}. This problem could be avoided by only including connected pairs in the sum, but this introduces a distortion for networks with a lot of unconnected pairs of nodes. These networks will then have a small average distance, which is only expected for networks with lots of connections \cite{F.Costa2007}. Instead, another definition of average distance can be used: the global efficiency $E$ \cite{F.Costa2007}
\begin{equation}
	E = \frac{1}{N(N-1)} \sum_{i \neq j} \frac{1}{d_{ij}} \ .
\end{equation}
This measurement quantifies the efficiency of the network in sending information between nodes \cite{F.Costa2007}. This assumes that the efficiency of sending information between two nodes is reciprocal to their distance. The harmonic mean of the geodesic distances is now defined as the reciprocal of the global efficiency \cite{F.Costa2007}
\begin{equation}
 	h = \frac{1}{E} \ .
\end{equation}
\newline
\textbf{Diameter and radius}\\
\newline
The diameter $diam(G)$ of a network $G$ is the maximum distance between any pair of nodes \cite{Goddard2010}. The eccentricity of a node is the maximum distance from that node to any other node. The radius $rad(G)$ is the minimum eccentricity among all nodes of $G$. Note that the diameter is the maximum eccentricity among all nodes in $G$ \cite{Goddard2010}.
\newpage
\noindent
\subsection{Clustering coefficient and transitivity}

@@

Let us denote the number of triangles as $N_{\Delta}$ and the number of connected triples as $N_3$. These two numbers can be obtained from the adjacency matrix \cite{F.Costa2007}

@@

It is not too hard to see that $N_{\Delta}(i)$ counts the number of edges between the neighbors of $i$ and that $N_3(i)$ is equal to $k_i(k_i-1)/2$, where $k_i$ is the degree of node $i$. It is then obvious that the local clustering coefficient $cc_i$ (Eq. (\ref{clus})) of a node and the transitivity $T_i$ (Eq. (\ref{trans})) of a node define the same quantity \cite{F.Costa2007}. In the case of the average clustering coefficient $\left<cc\right>$ and the global transitivity $T$ this is, however, not the case. The difference between the two definitions is that the average of Eq. (\ref{globalTrans}) gives the same weight to each triangle in the network, whereas Eq. (\ref{avClus}) gives the same weight to each node \cite{Newman2003}. This may lead to slightly different values since nodes with a higher degree may possibly be involved in a higher number of triangles than nodes with a lower degree \cite{F.Costa2007}.

@@

The modularity of a network with $N$ nodes and $L$ edges and partitioned into $n_c$ communities is calculated as follows. Each community has $N_c$ nodes and $L_c$ edges ($c = 1,..., n_c$) \cite{Albert2016}. The modularity of a community $Q_c$ is then calculated by measuring the difference between the network's real wiring diagram (which is given by the adjacency matrix $A_{ij}$) and the expected number of edges between $i$ and $j$ if the network is randomly wired (where $i$ and $j$ are two nodes that lie in the same community $C_c$) \cite{Albert2016}. This last quantity is given by $p_{ij}$, which is obtained by randomizing the original network while keeping the expected degree of each node unchanged \cite{Albert2016}. Hence, the modularity of a community $Q_c$ becomes \cite{Albert2016}

@@

where the factor $\frac{1}{2L}$ is a normalization factor. Using \cite{Albert2016}

@@

Modularity is often used in community detection algorithms. It is thus important to be aware of the limitations of modularity. One of the biggest shortcomings of modularity maximization as community detection is its resolution limit. Modularity maximization forces small communities into larger ones \cite{Albert2016}. If two communities $A$ and $B$ are merged, the modularity changes by \cite{Albert2016}

@@

and if there is at least one edge between the two communities; even if $A$ and $B$ are distinct communities \cite{Albert2016}! Eq. (\ref{reslim}) is called the resolution limit: communities that are smaller than this limit will not be detected by modularity maximization \cite{Albert2016}.

@@

	P(k) = \binom{N-1}{k}p^k (1-p)^{N-1-k} \ .

@@

	\left<k\right> &= \sum_{k=0}^{N-1} k\  P(k)  \\

@@

	P(k) = \mathrm{e}^{-\left<k\right>} \frac{\left<k\right>^k}{k!} \ .

@@

This can be easily understood since choosing three nodes out of $N$ is given by $\binom{N}{3}$. The chance that they are connected by three edges is given by $p^3$, which gives the number of triangles; the chance that the are connected by two edges is $p^2$, which determines the number of connected triples (recall that this model creates random networks with no self-loops and no edge duplicates). If the average degree $\left<k\right>$ is constant, the average clustering constant goes as $\left<cc\right> \sim \mathcal{O}(1/N)$, which becomes small for large networks \cite{Clauset2011}. \\

@@

A more precise calculation can be found in Appendix \ref{ER}, Section \ref{ERdist}. Eq. (\ref{diameter}) might, however, be a better approximation to the average distance $\left<d\right>$ between two randomly chosen nodes \cite{Albert2014}. This is because $d_{\textrm{max}}$ is often dominated by a few extreme paths, whereas $\left<d\right>$ is averaged over all node pairs and thus suppresses these fluctuations \cite{Albert2014}. Hence, the average distance of a random network is defined by \cite{Albert2014}

@@

To summarize: the random Erd\H{o}s-R\'{e}nyi network has a bell shaped degree distribution which approximates a Poisson distribution for $N \gg 1$; the average clustering coefficient becomes small for large networks and the average distance depends logarithmic on the system sizes. The Erd\H{o}s-R\'{e}nyi network thus possesses one of the two properties of the ``small world phenomenon'', a short average distance.\\
